---
title: compare() two tables
author: zchmielewska
date: '2020-03-20'
slug: compare-two-tables
categories: []
tags: []
description: ''
---

# Usage 

A big chunk of actuarial life consists of comparing stuff.

After preparation of assumptions tables and policy data for valuation models,
it's useful to check how they change over a reporting period 
to understand the magnitude of movements.
During the model development, it's important to understand the impact
of each of the components on the final results.

The `compare()` function helps to see the differences of two tables.

```{r}
library("actuaryr")
```

```{r}
x <- data.frame(
  v1 = c(1, 2, 3),
  v2 = c("a", "b", "c"),
  v4 = c(10, 10, 10),
  stringsAsFactors = FALSE
)
print(x)
```

```{r}
y <- data.frame(
  v1 = c(1, 2, 4, 5),
  v2 = c("a", "b", "d", "e"),
  v3 = c(6, 7, 8, 9),
  v4 = c("10", "10", "10", "10"),
  stringsAsFactors = FALSE
)
print(y)
```

```{r}
compare(x, y)
```

The result of the comparison is:

* for numericals - the absolute difference,
* for characters - either `the_same` or `different`.

The function prints out diagnostic messages. 
These diagnostic messages result from the approach taken for the comparison. 
The approach is described in the next section.

# Comparison approach

`compare()` uses the following approach:

1. Only common columns are compared. 
2. Only common rows are compared. 
3. Columns are coerced to common types.

Let's see some examples of this behaviour.

## Example 1 - common columns

Only common columns of the two data frames are compared. 
The rest of the columns is ignored for the comparison.

```{r echo = FALSE}
x <- data.frame(
  v1 = c(1, 2, 3),
  v2 = c("a", "b", "c"),
  stringsAsFactors = FALSE
)

y <- data.frame(
  v1 = c(1, 2, 2),
  v3 = c("d", "e", "f"),
  stringsAsFactors = FALSE
)
```

```{r}
print(x)
```

```{r}
print(y)
```

```{r}
compare(x, y)
```

Only `v1` is a common column between the two tables, 
so the rest of the columns (`v2` in LHS and `v3` in RHS) has been ignored.

## Example 2 - common rows

Only common rows of the two data frames are compared. 
If one data frame has more rows, the extra rows are removed for the comparison.

```{r echo = FALSE}
x <- data.frame(
  v1 = c(1, 2, 3),
  v2 = c("a", "b", "c"),
  stringsAsFactors = FALSE
)

y <- data.frame(
  v1 = c(1, 2),
  v2 = c("a", "c"),
  stringsAsFactors = FALSE
)
```

```{r}
print(x)
```

```{r}
print(y)
```

```{r}
compare(x, y)
```

The diagnostic message concerns the issue that 
there are 3 rows in LHS and only 2 rows in RHS.
The 3<sup>rd</sup> row in LHS has been removed for the comparison.
Only the first 2 rows are compared.

## Example 3 - common types

If the same columns have different types between the two tables, 
the weaker type gets coerced to the stronger type. 

Strength order: logical < integer < double < character.

```{r echo = FALSE}
x <- data.frame(
  v1 = c(1, 2, 3)
)

y <- data.frame(
  v1 = c("1", "2", "3"),
  stringsAsFactors = FALSE
)
```

```{r}
str(x)
```

```{r}
str(y)
```

```{r}
compare(x, y)
```

The `v1` column in LHS is a double and in RHS a character.
For comparison, the `v1` column LHS is coerced to a character.