---
title: Change an interest rates curve
author: zchmielewska
date: '2020-04-26'
slug: change-period-and-change-type-of-an-interest-rate-curve
categories: []
tags: []
description: ''
---
# Load package

To load `actuaryr` package, use:
```{r}
library(actuaryr)
```

# Change type of an interest rates curve

The two common types of interest rates curve are **forward** and **spot**.
With **forward** rates, we move along the timeline period by period.
With **spot** rates, we jump directly from the beginning of the period 
to the point on a timeline.

Imagine an interest rates curve with yearly forward rates: 1.8%, 2%, 2.2%.


```{r}
irc_forward <- as_ircurve(rates = c(0.018, 0.02, 0.022))
```

The default arguments' values for `as_ircurve()` are 
`type = "forward"` and `period = "year"`.

```{r}
print(irc_forward)
```

The value of `100` after 3 years amounts to 106.12.

```{r}
100 * prod(1 + irc_forward)
```

The **forward** rates can be changed to **spot** rates with `change_type()`.

```{r}
irc_spot <- change_type(irc_forward, to = "spot")
```

The value of 100 after 3 years using **spot** rates amounts to 106.12 as well.

```{r}
100 * (1 + irc_spot[3])^3
```

The process of changing the curve's type is *recursive*. 
If you change the type from **forward** to **spot** 
and then from **spot** to **forward**, you'll receive the same curve.

```{r, warning = FALSE}
# Let's import the pipe (%>%) from magrittr
library(magrittr)

ircurve1 <- as_ircurve(c(0.1, 0.2, 0.3))
ircurve2 <- ircurve1 %>% change_type(to = "spot") %>% change_type(to = "forward")
print(ircurve2)

```

# Change period of an interest rate curve

The interest rates are usually presented as **yearly** rates. 
Yearly rates tell us how much will the money be worth in a year's time.
It's also convenient to use **monthly** interest rates, because
we receive our paychecks on a monthly basis and pay bills on a monthly basis
as well.

You can change the interest rates curve from yearly to monthly and 
vice versa with `change_period()`.

## From month to year

You can change the monthly rates in to yearly rates with 
`change_period(ircurve, to = "year").

```{r}
irc_month <- as_ircurve(rates = c(rep(0.001, 12), rep(0.002, 12)), period = "month")
irc_year  <- change_period(irc_month, to = "year")
```

With monthly rates, you gain 3.66 on 100 after two years.

```{r}
# Change in value in two years using monthly rates
100 * prod(1 + irc_month)
```

With yearly rates, too!

```{r}
# Change in value in one year using yearly rate
100 * prod(1 + irc_year)
```

If the monthly curve does not cover the whole year, the remaining part is filled
with **the last** monthly rate.

```{r, warning = FALSE}
# This transformation...
rates1     <- seq(0.001, 0.009, 0.001)
irc_month1 <- as_ircurve(rates1, period = "month")
irc_year1  <- change_period(irc_month1, to = "year")

# ...is equivalent to this transformation.
rates2     <- c(seq(0.001, 0.009, 0.001), rep(0.009, 3))
irc_month2 <- as_ircurve(rates2, period = "month")
irc_year2  <- change_period(irc_month2, to = "year")
```

The change from **year** to **month** assumes that monthly rates within a year
are equal.

```{r}
irc_year <- as_ircurve(rates = (1+0.01)^12-1)
change_period(irc_year, to = "month")
```

Hence, the transformation from month to year and from year to month
is **not** recursive, unless monthly rates are equal.

```{r}
# This curve...
rates      <- seq(0.001, 0.012, 0.001)
irc_month1 <- as_ircurve(rates, period = "month")

# ...is different than this curve
irc_month2 <- irc_month1 %>% change_period(to = "year") %>% change_period(to = "month")
```

